Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models

### 2020-06-11 analysis/paper/paper.Rmd
revisions in Discussion of paper.Rmd

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act (ESA) delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat restoration actions to recover depleted salmon populations. Moreover, the models presented here can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life stage specific habitat is limiting. For example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required for ESA delisting, but that juvenile capacity may not be sufficient to support those levels of adult spawning. In such a case, habitat restoration actions may be most cost-effectively and sustainably directed towards juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions.

@@

When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999]. However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults. Such datasets are not available everywhere, nor do they necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale. QRF models require a large number of paired fish-habitat sites, but if done across a wide variety of habitat types this data could theoretically be collected in a single season. Or the dataset could be built over time, sampling different locations in different years. If the latter approach is taken, some thought must be given to possible year effects and how to best deal with them.

The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds goes a long way to justifying this approach. Although built from completely different data, when these mulitple lines of evidence converge it lends credence to the QRF results.

@@

The results of the QRF parr capacity model for spring/summer Chinook salmon meet many biological expectations. Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width-to-depth ratio grow, when temperatures are cooler [@Bjornn1991; @Raleigh1986; @Brett1952], when there is less fine sediment [@Allen2000; @Bjornn1991; @Hillman1987], when there is more fish cover [@Holecek2009; @Bjornn1991; @Hillman1987], when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for streambed complexity) increases. These are all patterns that emerged from the fish-habitat data, and where available, match those fish-habitat relationships identified qualitatively in other studies [@Mossop2006].

The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response. In many ways, these metrics define habitat quantity. However, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sendiment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.

Moreover, predictions of available habitat capacity seemed to pass the "eye test" when comparing predictions to the actual physical habitat at CHaMP survey locations. That is, areas with higher predicted capacities had habitat that, in our opinion, looked like habitat that would support more juvenile Chinook salmon, whereas areas with lower predicted capacities looked like poorer rearing habitat. As examples, Figure \@ref(fig:site-photo-high) is a photograph of a CHaMP survey site where the model predicted high capacity and is a site with an increases width-to-depth ratio and fish cover (in the form of mature willow and cottonwood riparian cover), contains pool-riffle sequences, and some variation in wetted depths. Alternatively, Figure \@ref(fig:site-photo-low) shows a site where the model predicted a much lower capacity, and visually, is much poorer rearing habitat for juvenile Chinook salmon e.g., lower wetted widths, discharge, width-to-depth ratio and standard deviation of depth. Pairing predictions of capacity from the QRF model with images and expert opinion of good versus bad habitat will be an important ongoing validation in addition to our comparisons with spawner-recruit models.

@@

Our extrapolation model was focused on extrapolating to other master sample points, because that is the dataset available to us, but the methodology could be improved. Extrapolating to reaches on a stream network, as opposed to points on the landscape, could improve the interpretability of the results. This would require a stream network with relevant attributes attached to each reach, similar to the GAAs we used. Another approach could be to move towards sampling habitat in a more spatially continuous fashion, covering most or all of a watershed, and builting a QRF model from that dataset. Even if the fish data was not done continously, estimates of capacity could be made directly from the QRF model across the entire stream network, without the need for an extrapolation model.

One of the potential downsides to the extrapolation approach used here is that the globally available attributes generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat. While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model. This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale. 
<!-- Do we need any more in this section? I commented out the paragraph below... -->

@@

Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River basin, there is a pressing need to develop new tools for habitat analyses.  Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015]. UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling. Advances in imaging techniques (e.g., multispectral imaging) and post processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].

Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly improve the broadscale application of QRF. Rapid advances drone technology further improves upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated data post-processing. All data collection efforts would be georeferenced and topologically compatible to increase repeatability of methods and data collection locations; a primary criticism of previous CHaMP survey efforts. The implementation of such a protocol would circumvent the need to extrapolate by collecting data for individual channel units in a rapid manner using remote sensing technologies, thereby reducing labor, providing a cost-effective tool for habitat data collection supporting status and trend evaluation and model products to better inform habitat restoration prioritization and planning.

@@

If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment. This understanding could be of crucial importance in the realm of conservation, when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally. We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity. As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey [@Sauer2017] to groundfish trawl surveys [@Bradburn2011]. The framework could also be applied to depleted, non-migratory, and isolated populations (e.g. desert pupfish *Cyprinodon macularius*) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed. Capacity estimates could also be used to potential translocation sites to determine if given sites can support an abundance considered viable before investing in translocation efforts.  ***other potential examples (coral reefs? turtles?)***

>>>>

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, the models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions.

The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds goes a long way to justifying this approach. Although built from completely different data, when these mulitple lines of evidence converge it lends credence to the QRF capacity prediction results.

@@

When it comes to estimating carrying capacity, spawner-recruit models are the gold standard [@Moussalli1986; @Myers1999]. However, such models require a long time-series of accurate estimates of adults and juveniles, with variation in the number of adults. Such datasets are not available everywhere, nor do they necessarily allow one to link capacity to habitat characteristics, except perhaps at the watershed scale. QRF models require a large number of paired fish-habitat sites, but if done across a wide variety of habitat types this data could theoretically be collected in a single (or very few) season(s). Or the dataset could be built over time, sampling different locations in different years. If the latter approach is taken, some thought must be given to possible year effects and how to best deal with them.

@@

The results of the QRF parr capacity model for Chinook salmon meet many biological expectations. Focusing on the partial dependence plots (Figure \@ref(fig:pdp-figure)), the QRF model predicts capacity to increase when the wetted width, discharge and the width:depth ratio grow, when temperatures are cooler [@Bjornn1991; @Raleigh1986; @Brett1952], when there is less fine sediment [@Allen2000; @Bjornn1991; @Hillman1987], when there is more fish cover [@Holecek2009; @Bjornn1991; @Hillman1987], when channel unit frequency increases and when the standard deviation of the wetted depth (a proxy for streambed complexity) increases. These are all patterns that emerged from the fish-habitat data, and where available, match those fish-habitat relationships identified qualitatively in other studies [@Mossop2006].

The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response. In many ways, these metrics define habitat quantity. However, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.

Moreover, predictions of available habitat capacity seemed to pass the "eye test" when comparing predictions to the actual physical habitat at CHaMP survey locations. That is, areas with higher predicted capacities had habitat that, in our opinion, looked like habitat that would support more juvenile Chinook salmon, whereas areas with lower predicted capacities looked like poorer rearing habitat. As examples, Figure \@ref(fig:site-photo-high) is a photograph of a CHaMP survey site where the model predicted high capacity and is a site with an increases width:depth ratio and fish cover (in the form of mature willow and cottonwood riparian cover), contains pool-riffle sequences, and some variation in wetted depths. Alternatively, Figure \@ref(fig:site-photo-low) shows a site where the model predicted a much lower capacity, and visually, is much poorer rearing habitat for juvenile Chinook salmon e.g., lower wetted widths, discharge, width:depth ratio and standard deviation of depth. Pairing predictions of capacity from the QRF model with images and expert opinion of good versus bad habitat will be an important ongoing validation in addition to our comparisons with spawner-recruit models.

@@

Our extrapolation model was focused on extrapolating to other master sample points, because that is the dataset available to us, but the methodology could be improved. Extrapolating to reaches on a stream network, as opposed to points on the landscape, could improve the interpretability of the results. This would require a stream network with relevant attributes attached to each reach, similar to the GAAs we used. Another approach could be to move towards sampling habitat in a more spatially continuous fashion, covering most or all of a watershed, and building a QRF model from that dataset. Even if the fish data was not done continuously, estimates of capacity could be made directly from the QRF model across the entire stream network, without the need for an extrapolation model.

One of the potential downsides to the extrapolation approach used here is that the globally available attributes generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat. While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model. This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale.

<!-- Do we need any more in this section? I commented out the paragraph below... -->

@@

Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River basin, there is a pressing need to develop new tools for habitat analyses. Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015]. UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling. Advances in imaging techniques (e.g., multispectral imaging) and post-processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].

Development of a standardized protocol to incorporate remotely sensed data (LiDAR, aerial imagery) into the collection of habitat metrics would greatly improve the broadscale application of QRF. Rapid advances in drone technology further improves upon traditional habitat data collection by leveraging 1) sub-meter global navigation satellite system (GNSS) receivers; 2) cost-effective drone imagery collection, image stitching, and photogrammetry; and 3) semi-automated data post-processing. All data collection efforts would be georeferenced and topologically compatible to increase repeatability of methods and data collection locations; a primary criticism of previous CHaMP survey efforts. The implementation of such a protocol would circumvent the need to extrapolate by collecting data for individual channel units in a rapid manner using remote sensing technologies, thereby reducing labor, providing a cost-effective tool for habitat data collection supporting status and trend evaluation and model products to better inform habitat rehabilitation prioritization and planning.

@@

If a species' carrying capacity is defined or constrained, at least in part, by the habitat in which it lives, then illuminating statistically how such habitat impacts carrying capacity can lead to understanding how a species interacts with its environment. This understanding could be of crucial importance in the realm of conservation, when dealing with an endangered or threatened species, but species/habitat interactions are a core element of ecological studies more generally. We have demonstrated how a quantile regression approach, coupled with a random forest framework, can be used to estimate these relationships, and predict an habitat's capacity. As large ecological datasets become more accessible, and the ability to measure large swaths of habitat more feasible, we believe this approach has many potential applications, from the North American breeding bird survey to groundfish trawl surveys. The framework could also be applied to depleted, non-migratory, and isolated populations (e.g. desert pupfish *Cyprinodon macularius*) to identify limiting factors in populations and/or determine whether a given habitat patch could support a viable population if limiting factors were addressed. Capacity estimates could also be used to evaluate potential translocation sites to determine if those sites could support an abundance considered viable before investing in translocation efforts.
